Modeling and analyzing the Corona-virus warning app with the Isabelle Infrastructure framework

### 2020-07-11 paper.tex
shrunk the paper to 16 pages

\newcommand{\symb}[1]{\makebox{\it #1}} 
\newcommand\ie{i.e.\!\,, }
\newcommand\rel{Re${\cal{L}}$}
%\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
%\newcommand*{\TODOfn}[2][noteC]{\TODO[#1]{[\footnote{\TODO[#1]{#2}}]}}
%\newcommand*{\TODOref}[2][todoC]{\TODOfn[#1]{\textbf{BIB-REF}: #2}}
\newcommand\ttand{\mbox{{$\land$}}}
\newcommand\ttor{\mbox{{$\lor$}}}
\newcommand\ttcap{\mbox{{$\cap$}}}
\newcommand\ttcup{\mbox{{$\cup$}}}
\newcommand\ttfun{\mbox{{$\Rightarrow$}}}
\newcommand\ttmimp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttimp{\mbox{{$\longrightarrow$}}}
\newcommand\ttequiv{\mbox{{$\equiv$}}}
\newcommand\ttexists{\mbox{{$\exists$}}}
\newcommand\ttforall{\mbox{{$\forall$}}}
\newcommand\ttneg{\mbox{{$\neg$}}}
\newcommand\ttneq{\mbox{{$\neq$}}}
\newcommand\ttin{\mbox{{$\in$}}}
\newcommand\ttnin{\mbox{{$\notin$}}}
\newcommand\ttImp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttlam{\mbox{\( \lambda \)}}
\newcommand\tttimes{\mbox{\( \times \)}}
\newcommand\ttlbrack{\mbox{\(\llbracket\)}}
\newcommand\ttrbrack{\mbox{\( \rrbracket \)}}
\newcommand\noie{\textit{i.e.},\xspace}
\newcommand\eg{~\textit{e.g.},\xspace}
\newcommand\noeg{\textit{e.g.},\xspace}
\newcommand\ttatI{\mbox{\( @_G \)}}
\newcommand\ttto[1]{\mbox{{$\to^{#1}$}}}
\newcommand\ttleq{\mbox{{$\le$}}}
\newcommand\ttrelI{\mbox{{$\to_{i}$}}}
\newcommand\ttrelIstar{\mbox{{$\to^*$}}}
\newcommand\ttrel[1]{\mbox{{$\to_{#1}$}}}
\newcommand\ttrelstar[1]{\mbox{{$\to_{#1}^*$}}}
\newcommand\ttalpha{\mbox{{$\alpha$}}}
\newcommand\ttsupseteq{\mbox{{$\supseteq$}}}
\newcommand\ttupdownarrow{\mbox{{$\Updownarrow$}}}
\newcommand\ttbigcap{\mbox{{$\bigcap$}}}
\newcommand\tttau{\mbox{{$\tau$}}}
\newcommand\ttsubseteq{\mbox{{$\subseteq$}}}
\newcommand\ttf{\mbox{{$f$}}}
\newcommand\ttvdash{\mbox{{$\vdash$}}}
\newcommand\ttref[1]{\mbox{\(\sqsubseteq_{#1}\)}}
\newcommand\ttrefV{\mbox{{$\sqsubseteq_V$}}}
\newcommand{\ttcalN}[1]{\mbox{{${\mathcal{N}}_{\texttt{#1}}$}}} 
\newcommand\ttNatt{\mbox{{$\mathcal{N}$}}}
\newcommand\ttattand[1]{\mbox{{$\oplus_{\wedge}^{#1}$}}}
\newcommand\ttattor[1]{\mbox{{$\oplus_{\vee}^{#1}$}}}
\newcommand\ttsigma{\mbox{{$\sigma$}}}
\newcommand\ttmref[1]{\mbox{{$\sqsubseteq_{#1}$}}}
\newcommand\ttmeref{\ttmref{\mathcal{E}}}
\newcommand\ttsn{\mbox{{\texttt{s}$_0$}}}
\newcommand\ttsigmap{\mbox{{$\sigma'$}}}
\newcommand\ttecal{\mbox{$\mathcal{E}$}}
%\newcommand\ttimg{\mbox{$\triangleleft$}}
\newcommand\ttimg{\mbox{\texttt{`}}}

@@

\section{Background and related work}
\label{sec:background}
\subsection{DP-3T and PEPP-PT}
\label{sec:history}
We are mainly concerned with the architecture and protocols proposed by the
DP-3T (\textit{Decentralized Privacy-Preserving Proximity Tracing}) project \cite{dp3t:github}.
The main reason to focus on this particular family of protocols is the
\textit{Exposure Notification Framework} (ENF) \cite{enf:proj}, jointly published by Apple and Google, that adopts
%the main
core principles of the DP-3T proposal. This API is not only used in the German
Corona-virus warning app but has the potential of being widely adopted in future
app developments that might emerge due to the reach of players like Apple and Google.

@@

All DP-3T protocols roughly work the same:
%\subsubsection{Basic DP-3T protocol: How does it work?}
%  How does it relate to ENF/CWA? (broad strokes)}
%
Upon installation, the app generates secret daily seeds to derive \textit{Ephemeral Ids}
(EphIDs) from them. EphIDs are generated locally with cryptographic methods and cannot be connected
to one another but only reconstructed from the secret seed they were derived from.

@@

\begin{figure}[htb]
  \begin{center}
    \includegraphics[width=5in]{DP-3T_basic_protocol}
  \end{center}
  \caption{Decentralized privacy-preserving proximity tracing protocol
    (as implemented by CWA)}
  \label{fig:dp3tprot}
\end{figure}

\subsection{Isabelle Infrastructure framework}
\label{sec:isainf}
%\TODO{Adapt: this is copied from FMBC paper}
The Isabelle Infrastructure is built in the interactive generic theorem prover
Isabelle/HOL \cite{npw:02}. As a framework, it supports formalization and proof of
systems with actors and policies. It originally emerged from verification of insider 
threat scenarios but it soon became clear that the theoretical concepts, like temporal
logic combined with Kripke structures and a generic notion of state transitions were
very suitable to be combined with attack trees into a formal security engineering process
\cite{suc:16} and framework \cite{kam:19a}.
%\subsubsection{Overview}
%\label{sec:isainfover}
Figure \ref{fig:theorystruc} gives an overview of the Isabelle Infrastructure 
framework with its layers of object-logics -- each level below embeds the one
above showing the novel contribution of this paper in blue on the top. 
\begin{figure}
\begin{center}
  \includegraphics[scale=.5]{theory_structure.pdf}
\end{center}
%\vspace{-.5cm}
\caption{Generic Isabelle Infrastructure framework applied to Corona-warning app.}
\label{fig:theorystruc}
\end{figure}

@@

The transition between states is triggered by non-parametrized
actions \texttt{get}, \texttt{move}, and \texttt{put} 
executed by actors. 
Actors are given by an abstract type \texttt{actor} and a function 
\texttt{Actor} that creates elements of that type from identities 
(of type \texttt{string} written \texttt{''s''} in Isabelle). 
Actors are contained in an infrastructure graph type \texttt{igraph}
constructed by its constructor \texttt{Lgraph}. 
\begin{ttbox}
{\bf datatype} igraph = 
         Lgraph (location \tttimes location)set 
                 location \ttfun identity set
                 identity \ttfun (string set \tttimes string set \tttimes efid)  
                 location \ttfun string \tttimes (dlm \tttimes data) set
                 location \ttfun efid set
                 actor \ttfun location \ttfun (identity \tttimes efid) set
\end{ttbox}
In the current application of the framework to the Corona-warning app case study, 
this graph contains a set of location pairs representing the %structure 
topology of the infrastructure
as a graph of nodes and a function%\footnote{ We use the common $\lambda$-abstraction, 
%e.g. \texttt{\ttlam x. True}, to define functions with parameters, here the 
%function returning True for any input $x$.} 
that assigns a set of actor identities to each node (location) in the graph.
The third component of an \texttt{igraph} assigns the credentials to each actor: 
a triple-valued function whose first range component is a set describing the credentials 
in the possession of an actor and the second component 
is a set defining the roles the actor can take on; 
most prominently the third component is the \texttt{efid} assigned to the actor. This is initially
just a natural number but will be refined to actually represent lists of ``Ephemeral'' Ids later
when refining the specification.
\begin{ttbox}
{\bf datatype} efid = Efid nat
\end{ttbox}
The fourth component of the type \texttt{igraph}
assigns security labelled data to locations, a feature not used in the
current application.
The second to last component assigns the set of efids of all currently present smart phones
to each location of the graph. The last component finally denotes the knowledge set of each
actor for each location: a set of pairs of actors and potential ids.

@@

The Privacy protection goal is to avoid deanonymisation. That is, an attacker should not be able to
disambiguate the set of pairs of real Ids and EphIds. This is abstractly expressed in the predicate
identifiable.
\begin{ttbox}
identifiable eid A \ttequiv is_singleton\{(Id,Eid). (Id, Eid) \ttin A \ttand Eid = eid\}
\end{ttbox}
The predicate identifiable is used to express as the global policy `Eve cannot deanonymize an Ephemeral
Id \texttt{eid} using the gathered knowledge':
%{\bf fixes} global_policy::[infrastructure, identity] \ttfun bool
%{\bf defines}  
\begin{ttbox}
  global_policy I eid \ttequiv
             \ttneg(identifiable eid 
                ((\ttbigcap (kgra(graphI I)(Actor ''Eve'')`(nodes(graphI I))))
                 - \{(x,y). x = ''Eve''\}))
\end{ttbox}

@@

For the DP-3T Ephemeral Ids \cite{dp3t:wp}, for each day $t$ a seed $SK_t$
is used to generate a list of length $n = 24 * 60 / L$, where $L$ is the duration for which
the Ephemeral Ids are posted by the smart phone
\begin{ttbox}
  EphId1 || ... || EphIdn = PRG(PRF(SKt,``broadcast key''))
\end{ttbox} 
``where PRF is a pseudo-random function (e.g., HMAC-SHA256), ``broadcast key'' is a
fixed, public string, and PRG is a pseudorandom generator (e.g. AES in counter mode)'' \cite{dp3t:wp}.

From a cryptographic point of view, the crucial properties of the Ephemeral Ids are that
they are purely random, therefore, they cannot be guessed, but at the same time if -- after
the actual encounter between sender and receiver -- the seed $SK_t$ is published, it is feasible
to relate any of the \texttt{EphIDi} to $SK_t$  for all \texttt{i} $\in \{1, \ldots, n\}$.
For a formalisation of this crucial cryptographic property in Isabelle it suffices to define a new type
of list of Ephemeral Ids \texttt{efidlist} containing the root $SK_t$ (the first \texttt{efid}),
a current \texttt{efid} indicated by a list pointer of type \texttt{nat}, and the actual list of
\texttt{efid}s.
\begin{ttbox}
{\bf{datatype}} efidlist = Efids "efid" "nat" "efid list"
\end{ttbox}
We define functions for this datatype: \texttt{efidsroot} returning the first of the three
consituents in an \texttt{efidlist} (the root $SK_t$); \texttt{efids\_index} giving the second
component, the index of the current \texttt{efid}; \texttt{efids\_inc\_ind} applied to an
\texttt{efidlist} increments the index; \texttt{efids\_cur} returning the current \texttt{efid}
from the list and \texttt{efids\_list} for the entire list (the third component).

The first step of refinement replaces the simple \texttt{efid} in the infrastructure type by the
new type \texttt{efidlist}. Note, that in the new datatype \texttt{igraph} this change affects only
the thrird component, the credentials. The last two componenets, the set of currently present efids,
and the knowledge set, still operate on the simple type \texttt{efid}.
\begin{ttbox}
{\bf datatype} igraph = 
         Lgraph (location \tttimes location)set 
                 location \ttfun identity set
                 identity \ttfun (string set \tttimes string set \tttimes efidlist)  
                 location \ttfun string \tttimes (dlm \tttimes data) set
                 location \ttfun efid set
                 actor \ttfun location \ttfun (identity \tttimes efid) set
\end{ttbox}
The refined state transition relation implements the possibility of changing the Ephemeral Ids
by the rule for the action \texttt{put}.
\begin{ttbox}
 {\bf{put}}: G = graphI I \ttImp  a \ttatI l \ttImp enables I l (Actor a) put \ttImp
      I' = Infrastructure (put_graph_efid A l (graphI I))(delta I)
 \ttImp I \ttrelI I' 
\end{ttbox}
The change to the infrastructure state is implemented in the function \texttt{put\_graph\_efid} that
increases the current index in the \texttt{efidlist} in the credential component \texttt{cgra g n} for
the ``putting'' actor identity \texttt{n} and inserts the current \texttt{efid} from that credential
component into the \texttt{egra} component, the set of currently ``present'' Ephemeral Ids at the location
\texttt{l}.
\begin{ttbox}
put_graph_efid n l g  \ttequiv
  Lgraph (gra g)(agra g)
         ((cgra g)(n := (credentials (cgra g n), roles (cgra g n),
                         efids_inc_ind(efemid (cgra g n)))))
         (lgra g)
         ((egra g)(l := insert (efids_cur(efemid (cgra g n)))(egra g l)))
         (kgra g)
\end{ttbox}
We can now apply the refinement by defining a datatype map from the refined infrastructure type
\texttt{InfrastructureOne.infrastructure} to the former one \texttt{Infrastructure.infrastructure}.
\begin{ttbox}
{\bf{definition}} refmap :: InfrastructureOne.infrastructure \ttfun
                          Infrastructure.infrastructure
{\bf{where}} refmap I =
  Infrastructure.Infrastructure 
    (Infrastructure.Lgraph
       (InfrastructureOne.gra (graphI I))
       (InfrastructureOne.agra (graphI I))
        (\ttlam h. repl_efr 
           (InfrastructureOne.cgra (graphI I)) h)
           (InfrastructureOne.lgra (graphI I))
           (InfrastructureOne.egra (graphI I))
           (\ttlam a. \ttlam l.
  (\ttlam (x,y).(x, efids_root(efemid(InfrastructureOne.cgra (graphI I) x))))
                       \ttimg(InfrastructureOne.kgra (graphI I)) a l))                                                    
\end{ttbox}  
This is then plugged into the parameter \texttt{\ttecal} of the refinement operator allowing to
prove \texttt{corona\_Kripke \ttref{\texttt{refmap}} corona\_KripkeO} where the latter is the
refined Kripke structure.

@@

\subsubsection{Second refinement step}
The persistent attack can be abbreviated informally by the action sequence \texttt{[get,move,move,get]}
performed by actors Eve, Bob, Eve, and Eve again, respectively.
How can a second refinement step avoid that Eve does the last \texttt{get} by imposing that after the
first \texttt{move} of Bob a \texttt{put} action must happen before Eve can do another \texttt{get}?
A very simple remedy to exclude this attack is to bind a \texttt{put} action after every \texttt{move}.
We can implement that change by a minimal update to the function \texttt{move\_graph\_a} (see Section \ref{sec:model})
by adding an increment (highlighted in the code snippet)
of the currently used Ephemeral Id before updating the \texttt{egre} component of the target location.
\begin{ttbox}
move\_graph\_a n l l' g \ttequiv ...
  (l' := insert (efids_cur({\bf efids\_inc\_ind}(efemid (cgra g n))))(egra g l))...
\end{ttbox}  
This is an action refinement because the move action is changed. It is a refinement, since
any trace of the refined model can still be mapped to a trace in the more abstract model just
omitting a few steps (the refinement relation is defined using the reflexive transitive closure
\texttt{\ttrelIstar}.

\section{Conclusions}
\label{sec:concl}
\subsection{Protection goals of attacks}
Bluetooth based proximity tracing gives rise to a variety of noteworthy privacy and security issues that deserve formal consideration (and verification). The debate emerging between advocates of centralized architectures and those in favor of a decentralized approach in particular yields a lot of interesting material detailing different attacks and possible mitigation strategies: \cite{pepppt:dp3tana}, \cite{pepppt:dp3tresp}, \cite{dp3t:psre}.

In terms of attack scenarios, connecting an Ephemeral Id to a particular device is the first step into multiple different attacks, we want to classify as deanonymisation attacks:
\begin{itemize}
\item \textbf{Tracking a device}
  \cite{dp3t:psre}[p9], \cite{pepppt:dp3tana}[p8]\\
  Bluetooth enabled devices introduce some fundamental risks, that are not
  specific to proximity detection: If the Bluetooth signal is linked to a fixed
  identifier, say a MAC address, constant broadcasting makes
  a device trackable.
  Periodically changing Ephemeral Ids can be an effective mitigation\footnote{%
    In addition Bluetooth has to be configured to use rotating MAC addresses
    instead of fixed ones, as is the default in modern smartphone operating
    systems. Also, rotation of Bluetooth identifiers has to happen
    in sync with MAC address randomization; the Apple/Google proposal
    \cite{enf:bluetooth} ensures this.}
  as they limit the time frame in which a certain id can be used to
  track the corresponding device, hence, rendering the attack not impossible
  but significantly less severe.
\item \textbf{Identify infected individuals}
  \cite{dp3t:psre}[p5], \cite{pepppt:dp3tana}[p9]\\
  Like any notification about, e.\,g., a STD, might inform a spouse about the
  infidelity of the other, given the right circumstances, being notified about a
  possible exposure to COVID-19 might reveal illicit information to
  the recipient like who it was that tested positive for the virus.
  This is an inherent vulnerability of any such system,
  whether it is analogous or digital (digitally aided).\footnote{
    As DP-3T emphasizes, it it, indeed, inherent to proximity tracing in general
    \cite{dp3t:psre}[p6].
    The likelihood of said circumstances to occur, however,
    might increase when notifications are carried out automatically.
    Therefore, we feel that it is, nonetheless, worth considering.}
  In contrast to tracking a Bluetooth enabled device,
  this attack is not considerably mitigated by changing Ephemeral Ids:
  %Once connected to a device or individual, an Ephemeral Id might yield the
  %desired information for up to 14 days
  A positive test result can be reported for as much as 14 days after the
  compromised Ephemeral Id was used and still reveal the desired information.
\end{itemize}

\begin{figure}[htb]
  \begin{center}
    \includegraphics[width=6in]{DP-3T_deanonymisation_attacks}
  \end{center}
  \caption{Two deanynomisation attacks: Tracking users and identifying confirmed COVID-19 cases}
  \label{fig:dp3tprot}
\end{figure}

\subsection{Summary and discussion of relevance of the approach}
We can establish in our formal framework an attack that even a system using changing
Ephemeral ids can be broken if the attacker physically follows a victim. This is a basic attack on
anonymity: a user's connection between his Ephemeral Ids and personal details (Iphone MAC or name)
is revealed to the attacker. The protection goal of privacy is thereby destroyed.

@@

Now, the refinement shows that although the Ephemeral Ids change regularly the same attack that has been
identified on the very abstract level (fixed ids) persists.
The refinement allows refining the datatype (Id $\mapsto$ EphId) but also delivers the usual trace
refinement (behaviours of the refined systems are a subset of the traces of the abstract system).
This persistence of the attack precisely shows which part of the system behaviour is responsible for the
attack. In a second refinement step using action refinement based on this insight from the (repeated)
attack, we can exclude the dangerous attack trace\footnote{Implication on the actual protection goals
  that are endangered by the attack on privacy are discussed in the larger context of the Corona-warning
  app (see previous Section)}.

>>>>

\newcommand{\symb}[1]{\makebox{\it #1}} 
\newcommand\ie{i.e.\!\,, }
\newcommand\rel{Re${\cal{L}}$}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand*{\TODOfn}[2][noteC]{\TODO[#1]{[\footnote{\TODO[#1]{#2}}]}}
\newcommand*{\TODOref}[2][todoC]{\TODOfn[#1]{\textbf{BIB-REF}: #2}}
\newcommand\ttand{\mbox{{$\land$}}}
\newcommand\ttor{\mbox{{$\lor$}}}
\newcommand\ttcap{\mbox{{$\cap$}}}
\newcommand\ttcup{\mbox{{$\cup$}}}
\newcommand\ttfun{\mbox{{$\Rightarrow$}}}
\newcommand\ttmimp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttimp{\mbox{{$\longrightarrow$}}}
\newcommand\ttequiv{\mbox{{$\equiv$}}}
\newcommand\ttexists{\mbox{{$\exists$}}}
\newcommand\ttforall{\mbox{{$\forall$}}}
\newcommand\ttneg{\mbox{{$\neg$}}}
\newcommand\ttneq{\mbox{{$\neq$}}}
\newcommand\ttin{\mbox{{$\in$}}}
\newcommand\ttnin{\mbox{{$\notin$}}}
\newcommand\ttImp{\mbox{{$\Longrightarrow$}}}
\newcommand\ttlam{\mbox{\( \lambda \)}}
\newcommand\tttimes{\mbox{\( \times \)}}
\newcommand\ttlbrack{\mbox{\(\llbracket\)}}
\newcommand\ttrbrack{\mbox{\( \rrbracket \)}}
\newcommand\noie{\textit{i.e.},\xspace}
\newcommand\eg{~\textit{e.g.},\xspace}
\newcommand\noeg{\textit{e.g.},\xspace}
\newcommand\ttatI{\mbox{\( @_G \)}}
\newcommand\ttto[1]{\mbox{{$\to^{#1}$}}}
\newcommand\ttleq{\mbox{{$\le$}}}
\newcommand\ttrelI{\mbox{{$\to_{i}$}}}
\newcommand\ttrelIstar{\mbox{{$\to^*$}}}
\newcommand\ttrel[1]{\mbox{{$\to_{#1}$}}}
\newcommand\ttrelstar[1]{\mbox{{$\to_{#1}^*$}}}
\newcommand\ttalpha{\mbox{{$\alpha$}}}
\newcommand\ttsupseteq{\mbox{{$\supseteq$}}}
\newcommand\ttupdownarrow{\mbox{{$\Updownarrow$}}}
\newcommand\ttbigcap{\mbox{{$\bigcap$}}}
\newcommand\tttau{\mbox{{$\tau$}}}
\newcommand\ttsubseteq{\mbox{{$\subseteq$}}}
\newcommand\ttf{\mbox{{$f$}}}
\newcommand\ttvdash{\mbox{{$\vdash$}}}
\newcommand\ttref[1]{\mbox{\(\sqsubseteq_{#1}\)}}
\newcommand\ttrefV{\mbox{{$\sqsubseteq_V$}}}
\newcommand{\ttcalN}[1]{\mbox{{${\mathcal{N}}_{\texttt{#1}}$}}} 
\newcommand\ttNatt{\mbox{{$\mathcal{N}$}}}
\newcommand\ttattand[1]{\mbox{{$\oplus_{\wedge}^{#1}$}}}
\newcommand\ttattor[1]{\mbox{{$\oplus_{\vee}^{#1}$}}}
\newcommand\ttsigma{\mbox{{$\sigma$}}}
\newcommand\ttmref[1]{\mbox{{$\sqsubseteq_{#1}$}}}
\newcommand\ttmeref{\ttmref{\mathcal{E}}}
\newcommand\ttsn{\mbox{{\texttt{s}$_0$}}}
\newcommand\ttsigmap{\mbox{{$\sigma'$}}}
\newcommand\ttecal{\mbox{$\mathcal{E}$}}
%\newcommand\ttimg{\mbox{$\triangleleft$}}
\newcommand\ttimg{\mbox{\texttt{`}}}

@@

\section{Background and related work}
\label{sec:background}
\subsection{DP-3T and PEPP-PT}
\label{sec:history}
We are mainly concerned with the architecture and protocols proposed by the
DP-3T (\textit{Decentralized Privacy-Preserving Proximity Tracing}) project \cite{dp3t:github}.
The main reason to focus on this particular family of protocols is the
\textit{Exposure Notification Framework} (ENF), jointly published by Apple and Google, that adopts
%the main
core principles of the DP-3T proposal. This API is not only used in the German
Corona-virus warning app but has the potential of being widely adopted in future
app developments that might emerge due to the reach of players like Apple and Google.

@@

All DP-3T protocols roughly work the same:
%\subsubsection{Basic DP-3T protocol: How does it work?}
%  How does it relate to ENF/CWA? (broad strokes)}
%
Upon installation, the app generates secret daily seeds to derive \textit{Ephemeral IDs}
(EphIDs) from them. EphIDs are generated locally with cryptographic methods and cannot be connected
to one another but only reconstructed from the secret seed they were derived from.

@@

\begin{figure}[htb]
\vspace{-.5cm}
  \begin{center}
    \includegraphics[width=5in]{DP-3T_basic_protocol}
  \end{center}
\vspace{-.5cm}
  \caption{Decentralized privacy-preserving proximity tracing protocol of CWA}
  \label{fig:dp3tprot}
\vspace{-.5cm}
\end{figure}

\subsection{Isabelle Infrastructure framework}
\label{sec:isainf}
%\TODO{Adapt: this is copied from FMBC paper}
The Isabelle Infrastructure is built in the interactive generic theorem prover
Isabelle/HOL \cite{npw:02}. As a framework, it supports formalization and proof of
systems with actors and policies. It originally emerged from verification of insider 
threat scenarios but it soon became clear that the theoretical concepts, like temporal
logic combined with Kripke structures and a generic notion of state transitions were
very suitable to be combined with attack trees into a formal security engineering process
\cite{suc:16} and framework \cite{kam:19a}.
%\subsubsection{Overview}
%\label{sec:isainfover}
Figure \ref{fig:theorystruc} gives an overview of the Isabelle Infrastructure 
framework with its layers of object-logics -- each level below embeds the one
above showing the novel contribution of this paper in blue on the top. 
\begin{figure}
\vspace{-.5cm}
  \begin{center}
  \includegraphics[scale=.35]{theory_structure.pdf}
\end{center}
\vspace{-.5cm}
\caption{Generic Isabelle Infrastructure framework applied to Corona-warning app.}
\label{fig:theorystruc}
\vspace{-.5cm}
\end{figure}

@@

The transition between states is triggered by non-parameterized
actions \texttt{get}, \texttt{move}, and \texttt{put} 
executed by actors. 
Actors are given by an abstract type \texttt{actor} and a function 
\texttt{Actor} that creates elements of that type from identities 
(of type \texttt{string} written \texttt{''s''} in Isabelle). 
Actors are contained in an infrastructure graph type \texttt{igraph}
constructed by its constructor \texttt{Lgraph}. 
\begin{ttbox}
{\bf datatype} igraph = 
         Lgraph (location \tttimes location)set 
                 location \ttfun identity set
                 identity \ttfun (string set \tttimes string set \tttimes efid)  
                 location \ttfun string \tttimes (dlm \tttimes data) set
                 location \ttfun efid set
                 actor \ttfun location \ttfun (identity \tttimes efid) set
\end{ttbox}
In the current application of the framework to the Corona-warning app case study, 
this graph contains a set of location pairs representing the %structure 
topology of the infrastructure
as a graph of nodes and a function%\footnote{ We use the common $\lambda$-abstraction, 
%e.g. \texttt{\ttlam x. True}, to define functions with parameters, here the 
%function returning True for any input $x$.} 
that assigns a set of actor identities to each node (location) in the graph.
The third component of an \texttt{igraph} assigns the credentials to each actor: 
a triple-valued function whose first range component is a set describing the credentials 
in the possession of an actor and the second component 
is a set defining the roles the actor can take on; 
most prominently the third component is the \texttt{efid} assigned to the actor. This is initially
just a natural number but will be refined to actually represent lists of ``Ephemeral'' Ids later
when refining the specification.
\begin{ttbox}
{\bf datatype} efid = Efid nat
\end{ttbox}
The fourth component of the type \texttt{igraph}
assigns security labeled data to locations, a feature not used in the
current application.
The second to last component assigns the set of efids of all currently present smart phones
to each location of the graph. The last component finally denotes the knowledge set of each
actor for each location: a set of pairs of actors and potential ids.

@@

The Privacy protection goal is to avoid deanonymization. That is, an attacker should not be able to
disambiguate the set of pairs of real Ids and EphIds. This is abstractly expressed in the predicate
identifiable.
\begin{ttbox}
identifiable eid A \ttequiv is_singleton\{(Id,Eid). (Id, Eid) \ttin A \ttand Eid = eid\}
\end{ttbox}
The predicate identifiable is used to express as the global policy `Eve cannot deanonymize an Ephemeral
Id \texttt{eid} using the gathered knowledge':
%{\bf fixes} global_policy::[infrastructure, identity] \ttfun bool
%{\bf defines}  
\begin{ttbox}
  global_policy I eid \ttequiv
             \ttneg(identifiable eid 
                ((\ttbigcap (kgra(graphI I)(Actor ''Eve'')`(nodes(graphI I))))
                 - \{(x,y). x = ''Eve''\}))
\end{ttbox}

@@

For the DP-3T Ephemeral Ids \cite{dp3t:wp}, for each day $t$ a seed $SK_t$
is used to generate a list of length $n = 24 * 60 / L$, where $L$ is the duration for which
the Ephemeral Ids are posted by the smart phone
\begin{ttbox}
  EphId1 || ... || EphIdn = PRG(PRF(SKt,``broadcast key''))
\end{ttbox} 
``where PRF is a pseudo-random function (e.g., HMAC-SHA256), ``broadcast key'' is a
fixed, public string, and PRG is a pseudo-random generator (e.g. AES in counter mode)'' \cite{dp3t:wp}
%\TODOref{DP-3T Whitepaper}.

From a cryptographic point of view, the crucial properties of the Ephemeral Ids are that
they are purely random, therefore, they cannot be guessed, but at the same time if -- after
the actual encounter between sender and receiver -- the seed $SK_t$ is published, it is feasible
to relate any of the \texttt{EphIDi} to $SK_t$  for all \texttt{i} $\in \{1, \ldots, n\}$.
For a formalization of this crucial cryptographic property in Isabelle it suffices to define a new type
of list of Ephemeral Ids \texttt{efidlist} containing the root $SK_t$ (the first \texttt{efid}),
a current \texttt{efid} indicated by a list pointer of type \texttt{nat}, and the actual list of
\texttt{efid}s.
\begin{ttbox}
{\bf{datatype}} efidlist = Efids "efid" "nat" "efid list"
\end{ttbox}
We define functions for this datatype: \texttt{efidsroot} returning the first of the three
constituents in an \texttt{efidlist} (the root $SK_t$); \texttt{efids\_index} giving the second
component, the index of the current \texttt{efid}; \texttt{efids\_inc\_ind} applied to an
\texttt{efidlist} increments the index; \texttt{efids\_cur} returning the current \texttt{efid}
from the list and \texttt{efids\_list} for the entire list (the third component).

The first step of refinement replaces the simple \texttt{efid} in the infrastructure type by the
new type \texttt{efidlist}. Note, that in the new datatype \texttt{igraph} this change affects only
the third component, the credentials to become
\begin{ttbox}
   identity \ttfun (string set \tttimes string set \tttimes efidlist)
\end{ttbox}  
The last two components, the set of currently present efids,
and the knowledge set, remain the same and still operating on the simple type \texttt{efid}.
%\begin{ttbox}
%{\bf datatype} igraph = 
%         Lgraph (location \tttimes location)set 
%                 location \ttfun identity set
%
%                 location \ttfun string \tttimes (dlm \tttimes data) set
%                 location \ttfun efid set
%                 actor \ttfun location \ttfun (identity \tttimes efid) set
%\end{ttbox}
The refined state transition relation implements the possibility of changing the Ephemeral Ids
by the rule for the action \texttt{put} that resembles very much the rule for \texttt{get}.
%\begin{ttbox}
% {\bf{put}}: G = graphI I \ttImp  a \ttatI l \ttImp enables I l (Actor a) put \ttImp
%      I' = Infrastructure (put_graph_efid A l (graphI I))(delta I)
% \ttImp I \ttrelI I' 
%\end{ttbox}
The important change to the infrastructure state is implemented in the function \texttt{put\_graph\_efid} that
increases the current index in the \texttt{efidlist} in the credential component \texttt{cgra g n} for
the ``putting'' actor identity \texttt{n} and inserts the current \texttt{efid} from that credential
component into the \texttt{egra} component, the set of currently ``present'' Ephemeral Ids at the location
\texttt{l}.
\begin{ttbox}
put_graph_efid n l g  \ttequiv
  Lgraph (gra g)(agra g)
         ((cgra g)(n := (credentials (cgra g n), roles (cgra g n),
                         efids_inc_ind(efemid (cgra g n)))))
         (lgra g)
         ((egra g)(l := insert (efids_cur(efemid (cgra g n)))(egra g l)))
         (kgra g)
\end{ttbox}
We can now apply the refinement by defining a datatype map from the refined infrastructure type
\texttt{InfrastructureOne.infrastructure} to the former one \texttt{Infrastructure.infrastructure}.
\begin{ttbox}
{\bf{definition}} refmap :: InfrastructureOne.infrastructure \ttfun
                          Infrastructure.infrastructure
{\bf{where}} refmap I =
  Infrastructure.Infrastructure 
    (Infrastructure.Lgraph
       (InfrastructureOne.gra (graphI I))
       (InfrastructureOne.agra (graphI I))
        (\ttlam h. repl_efr 
           (InfrastructureOne.cgra (graphI I)) h)
           (InfrastructureOne.lgra (graphI I))
           (InfrastructureOne.egra (graphI I))
           (\ttlam a. \ttlam l.
  (\ttlam (x,y).(x, efids_root(efemid(InfrastructureOne.cgra (graphI I) x))))
                       \ttimg(InfrastructureOne.kgra (graphI I)) a l))                                                    
\end{ttbox}  
This is then plugged into the parameter \texttt{\ttecal} of the refinement operator allowing to
prove \texttt{corona\_Kripke \ttref{\texttt{refmap}} corona\_KripkeO} where the latter is the
refined Kripke structure.

@@

\subsubsection{Second refinement step}
The persistent attack can be abbreviated informally by the action sequence \texttt{[get,move,move,get]}
performed by actors Eve, Bob, Eve, and Eve again, respectively.
How can a second refinement step avoid that Eve does the last \texttt{get} by imposing that after the
first \texttt{move} of Bob a \texttt{put} action must happen before Eve can do another \texttt{get}?
A very simple remedy to exclude this attack is to bind a \texttt{put} action after every \texttt{move}.
We can implement that change by a minimal update to the function \texttt{move\_graph\_a} (see Section \ref{sec:model})
by adding an increment (highlighted in the code snippet)
of the currently used Ephemeral Id before updating the \texttt{egra} component of the target location.
\begin{ttbox}
move\_graph\_a n l l' g \ttequiv ...
  (l' := insert (efids_cur({\bf efids\_inc\_ind}(efemid (cgra g n))))(egra g l))...
\end{ttbox}  
This is an action refinement because the move action is changed. It is a refinement, since
any trace of the refined model can still be mapped to a trace in the more abstract model just
omitting a few steps (the refinement relation is defined using the reflexive transitive closure
\texttt{\ttrelIstar}.

\section{Summary and discussion of relevance of the approach}
%{Conclusions}
\label{sec:concl}
%\subsection{Protection goals of attacks}
%\subsection
We can establish in our formal framework an attack that even a system using changing
Ephemeral ids can be broken if the attacker physically follows a victim. This is a basic attack on
anonymity: a user's connection between his Ephemeral Ids and personal details (Iphone MAC or name)
is revealed to the attacker. The protection goal of privacy is thereby destroyed.

@@

Now, the refinement shows that although the Ephemeral Ids change regularly the same attack that has been
identified on the very abstract level (fixed ids) persists.
The refinement allows refining the datatype (Id $\mapsto$ EphId) but also delivers the usual trace
refinement (behaviours of the refined systems are a subset of the traces of the abstract system).
This persistence of the attack precisely shows which part of the system behaviour is responsible for the
attack. In a second refinement step using action refinement based on this insight from the (repeated)
attack, we can exclude the dangerous attack trace.
%\footnote{Implication on the actual protection goals
%  that are endangered by the attack on privacy are discussed in the larger context of the Corona-warning
%  app (see previous Section)}.
